Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 1

### 2024-09-16 main.tex
First full draft of the paper done, V1.

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage[natbib=true,style=authoryear]{biblatex}
% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\addbibresource{references.bib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage{todonotes}

@@

\title{Robotic Bicycle Balance Assistance Reduces Probability of Falling at Low
Speeds When Subjected to Mechanical Perturbations}

@@

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add a
  controlled steering motor to a consumer bicycle that stabilizes the bicycle
  at low speeds. To test the motor's assistance during falls, we apply varying
  magnitude external handlebar perturbations to the bicycle while ridden on a
  treadmill with the balance assist system activated and deactivated.  The
  probability of not recovering from a handlebar perturbation decreases when
  the balance assist is activated at a speed of 6~\si{\kph}.
}

@@

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handledbar, which can relatively easily cause
a rider to fall.

@@

At the initiation of each perturbation we log the instananeous steer and roll
angles to characterize the configuration of the bicycle when perturbed. The
gain setting on the balance assist controller indicates if the assistance is
off \(g=0\) or on at two different levels: low \(g=8\) or high \(g=10\). A
recovery from the perturbation is successful if the person neither places their
foot down onto the treadmill surface or their wheel of the bicycle exits the
width of the treadmill belt. We record this as a binary variable \(f\) for
"fall". All measured variables are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering, as recommended by \cite{Enders2007}, with clusters being an
individual subject. The clusters are chosend as all data from an individual
subject because we are only interested in the association between the state of
the balance-assist system and the outcome of the perturbation. We expected
there to be a variation between participants in how well they are able to
resist the perturbation. However, this was not true. Cluster-mean centering
shows there to be no variation between participants. This fact allows us to
utilize a simple single-level logistic regression model, instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
set of data.

@@

Turning the balance-assist system on significantly \((p<0.05)\) reduces the
odds that a pertubation results in a fall while cycling at a speed of
6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
for the balance assist state on and off while keeping all other explanatory
factors at their centered mean values. This figure is created by setting all
explanatory variables to there mean and calculating the probality from
Eq~\ref{eq:log-regress} for only change in angular impulse given the estimates
in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance-assist
system halves (0.53) the odds that a perturbation results in a fall. This
figure shows that for relatively large impulses the probability to fall is
unity for both states of balance assist on and off. And for relatively small
impulses the probability to fall is null for both states. But for impulses in
the magnitude region (-1 to 0.5 STD), i.e. around the mean-centered fall
threshold, the probability of falling is significantly lowered with the balance
assist system on. The skewedness of the probaility curves arrives from the
interaction effects.\todo{check this statement about skewedness this and think
about it} Figure~\ref{fig:probability-10kph} shows the same result for the
10~\si{\kph} trials which has a similar trend of reducing the probability to
fall with the balance assist system turned on, but the effect is not
significant.
%
\begin{figure}
  \centering
  % NOTE : We could uncomment the longer subcaption if titles removed from
  % figures. It is currently redundant.
  \subcaptionbox{%Fall probability for the 6~\si{\kph} trials.
    \label{fig:probability-6kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{%Fall probability for the 10~\si{\kph} trials.
    \label{fig:probability-10kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance-assist on
    (blue) or off (red) when all other predictor variables are fixed at zero,
    except for the interaction effect between the balance-assist and the
    angular impulse. The ordinate axis is standard deviation around the mean
    (all perturbations), centered per participant.}
  \label{fig:probability}
\end{figure}

\section{Discussion}
%
We have shown that at 6~\si{\kph} the addition of balance assist control
reduces the chance that a rider will fall when perturbed around the limits of their
control authority. But this effect diminishes at the higher speed
scenario of 10~\si{\kph}. We were only able to test these two speed-gain
scenarios for mostly homogeneous sets of riders within the resources of this
research project, but additional experimental work could help understand more
completely the range and limits of the positive effect of the balance system.
For example, it is possible that simply increasing the controller gain at
10~\si{\kph} also results in a significant positive effect.

\subsection{Interpretation of the Results}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain
some insight.
But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall:
\(e^{\alpha_k}=0.53\).
This means if the odds are a 1000:1, turning on the balance-assist system
reduces the odds to 500:1.
However, in that case the probability that a fall occurs is only reduced from
$0.999$ to $0.998$.
If the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance-assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse.
This means that the balance assist is most effective for perturbations close to
the subject's personal threshold between falling or recovery and that large
perturbations will make you fall regardless of the balance assist's help.

To illustrate the effect of the balance-assist system on fall probability, we
will give an example of how the data collected during the experiments is used
to predict fall probability. We use \ref{eq:log-regress} and the coefficients
in \ref{tab:freq-coefs-6}. For simplicities sake, the interaction effects are
not included.  Let's assume that the mean angular impulse $\bar{L}$ of all the
perturbations applied to a participant is 100~\si{\newton}, and the standard
deviation $\sigma^{L}=15$. The centred and scaled angular impulse can be
calculated by subtracting $\bar{L}$ from the applied angular impulse $L$, and
dividing this by $\sigma^{L}$. The same applies for the perturbation order $j$,
initial roll angle $\phi_0$, and initial steer angle $\delta_0$.  If we take
the coefficients estimated for cycling at 6~\si{\kph}, the log-odds of falling
can be calculated as follows.
%
\begin{align}
  \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
  =
  \beta + \sum_{k}^{k=0}\alpha_k \frac{x_{ij}^{k}-\bar{x_{ij}^{k}}}{\sigma^{x^{k}}}
  =
  & -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}}
    -0.77\cdot\frac{10-20}{11.54}  \\
  & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot s \\
  =
  & 1.42 -0.64 \cdot s
\end{align}

The state of the balance-assist $s$ is a binary variable. If the balance-assist
is turned on, the log-odds that a fall occurs are decreased by 0.64. The odds
and probability can be calculated:

\begin{align}
  \frac{p_{ij}}{1-p_{ij}} = e^{1.42-0.64s} = e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53s
\end{align}

\begin{align}
  p_{ij}^{s=0} = \frac{4.14}{1 + 4.14} = 0.81
\end{align}

\begin{align}
  p_{ij}^{s=1} = \frac{4.14\cdot0.53}{1 + 4.14\cdot{0.53}]} = 0.69
\end{align}

Turning on the balance-assist system reduces the probability that the
perturbation results in a fall from 0.81 to 0.69.

@@

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why balance assist system did not have a statistically
significant effect at 10~\si{\kph}. As a bicycle travels at higher speeds, the
same perturbation magnitude causes larger lateral deviations. At 10~\si{\kph}
almost all falls were due to the bicycle exiting the maximum width of the
treadmill. If the same experiment was performed on an infinitely wide plane,
the riders may have recovered from more perturbations. At 6~\si{\kph} the
riders could often recover in the allotted treadmill width due to the smaller
lateral deviations. Our results are very much dependent on the two modes of
falling with use: exit the treadmill width or foot is placed on the belt. Cycle
paths are a similar width as the treadmill, so rider's are often limited in
width when recovering from a fall.

@@

Roll and steer angle are not a significant predictor of fall probability,
neither at 6 or at 10~\si{\kph}. We expected this to have an effect. If you are
in a rolled and steered state that is far from the upright equilibrium, then a
perturbation that further pushes you from the equilibrium should have some
additive or multiplicative effect on the resulting motion trajectory.

@@

\section{Conclusion}
%
TODO

>>>>

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage[natbib=true,style=authoryear]{biblatex}
% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\addbibresource{references.bib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage[disable]{todonotes}

@@

\title{Automatic Bicycle Balance Assistance Reduces Probability of Falling at
Low Speeds When Subjected to Handlebar Perturbations\\Version 1}

@@

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add a
  controlled steering motor to a consumer electric bicycle that stabilizes it
  at low speeds. To test the motor's assistance during falls, we apply varying
  magnitude external handlebar perturbations to the bicycle while ridden on a
  treadmill with the balance assist system activated and deactivated. The
  probability of not recovering from a handlebar perturbation decreases when
  the balance assist is activated at a travel speed of 6~\si{\kph}. Use of a
  balance assist system in real world bicycling will reduce the number of falls
  that occur near riders' control authority limits.
}

@@

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can relatively easily cause
a rider to fall.

@@

At the initiation of each perturbation we log the instantaneous steer and roll
angles to characterize the configuration of the bicycle when perturbed. The
gain setting on the balance assist controller indicates if the assistance is
off \(g=0\) or on at two different levels: low \(g=8\) or high \(g=10\). A
recovery from the perturbation is successful if the person neither places their
foot down onto the treadmill surface or their wheel of the bicycle exits the
width of the treadmill belt. We record this as a binary variable \(f\) for
"fall". All measured variables are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering, as recommended by \cite{Enders2007}, with clusters being an
individual subject. The clusters are chosen as all data from an individual
subject because we are only interested in the association between the state of
the balance-assist system and the outcome of the perturbation. We expected
there to be a variation between participants in how well they are able to
resist the perturbation. However, this was not true. Cluster-mean centering
shows there to be no variation between participants. This fact allows us to
utilize a simple single-level logistic regression model, instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
set of data.

@@

Turning the balance-assist system on significantly \((p<0.05)\) reduces the
odds that a perturbation results in a fall while cycling at a speed of
6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
for the balance assist state on and off while keeping all other explanatory
factors at their centered mean values. This figure is created by setting all
explanatory variables to there mean and calculating the probability from
Eq~\ref{eq:log-regress} for only change in angular impulse given the estimates
in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance-assist
system halves (0.53) the odds that a perturbation results in a fall. This
figure shows that for relatively large impulses the probability to fall is
unity for both states of balance assist on and off. And for relatively small
impulses the probability to fall is null for both states. But for impulses in
the magnitude region (-1 to 0.5 STD), i.e. around the mean-centered fall
threshold, the probability of falling is significantly lowered with the balance
assist system on. The skewness of the probability curves arrives from the
interaction effects.\todo{check this statement about skewedness this and think
about it} Figure~\ref{fig:probability-10kph} shows the same result for the
10~\si{\kph} trials which has a similar trend of reducing the probability to
fall with the balance assist system turned on, but the effect is not
significant.
%
\begin{figure}
  \centering
  % NOTE : We could uncomment the longer subcaption if titles removed from
  % figures. It is currently redundant.
  \subcaptionbox{%Fall probability for the 6~\si{\kph} trials.
    \label{fig:probability-6kph}}{
    \includegraphics[width=120mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{%Fall probability for the 10~\si{\kph} trials.
    \label{fig:probability-10kph}}{
    \includegraphics[width=120mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance-assist on
    (blue) or off (red) when all other predictor variables are fixed at zero,
    except for the interaction effect between the balance-assist and the
    angular impulse. The ordinate axis is standard deviation around the mean
    (all perturbations), centered per participant.}
  \label{fig:probability}
\end{figure}

\section{Discussion}
%
We have shown that at a 6~\si{\kph} riding speed the addition of balance assist
control reduces the chance that a rider will fall when perturbed around the
limits of their control authority. But this effect diminishes just below
significance at the higher speed scenario of 10~\si{\kph}. We were only able to
test these two speed-gain scenarios for mostly homogeneous sets of riders
within the resources of this research project, but additional experimental work
could help understand more completely the range and limits of the positive
effect of the balance system. For example, it is possible that simply
increasing the controller gain at 10~\si{\kph} also results in a significant
positive effect. A longitudinal study of normal use of the balance assist
bicycle compared to a control group could provide the strongest evidence of any
benefit we have seen in this more narrow scenario.

@@

\subsection{Application of the Logistic Model}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain some
insight.
But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall:
\(e^{\alpha_k}=0.53\).
This means if the odds are a 1000:1, turning on the balance-assist system
reduces the odds to 500:1.
However, that case the probability that a fall occurs is only reduced from
$0.999$ to $0.998$.
If the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance-assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse the rider was subjected to.
This means that the balance assist is most effective for perturbations close to
the subject's personal threshold between falling or recovery and that large
perturbations will make you fall regardless of the balance assist's help.

To illustrate the effect of the balance-assist system on fall probability, we
will give an example of how the data collected during the experiments is used
to predict fall probability. We use Equation~\ref{eq:log-regress} and the coefficients
in \ref{tab:freq-coefs-6}. For simplicity's sake, the interaction effects are
not included. For example, assuming that the mean angular impulse $\bar{L}$ of all the
perturbations applied to a participant is 100~\si{\newton}, and the standard
deviation $\sigma^{L}=15$. The centered and scaled angular impulse can be
calculated by subtracting $\bar{L}$ from the applied angular impulse $L$, and
dividing this by $\sigma^{L}$. The same applies for the perturbation order $j$,
initial roll angle $\phi_0$, and initial steer angle $\delta_0$.  If we take
the coefficients estimated for cycling at 6~\si{\kph}, the log-odds of falling
can be calculated as follows.
%
\begin{equation}
  \begin{split}
    \log \left( \frac{p_{ij}}{1-p_{ij}} \right)
    = & \quad
    \beta + \sum_{k}^{k=0}\alpha_k \frac{x_{ij}^{k}-\bar{x_{ij}^{k}}}{\sigma^{x^{k}}} \\
    = &
   -0.29 + 1.69\cdot\frac{110~\si{\newton} - 115\si{\newton}}{15\si{\newton}} -0.77\cdot\frac{10-20}{11.54}  \\
    & -0.25\cdot\frac{-6\si{\degree}-2\si{\degree}}{10\si{\degree}}
    -0.14\cdot\frac{1\si{\degree}+3\si{\degree}}{5\si{\degree}} -0.64\cdot c \\
    = &
    1.42 -0.64 \cdot c
  \end{split}
\end{equation}
%
The state of the balance-assist \(s\) is a binary variable. If the
balance-assist is turned on, the log-odds that a fall occurs are decreased by
0.64. The odds and probability can be calculated:
%
\begin{align}
  \frac{p_{ij}}{1-p_{ij}} & = e^{1.42-0.64c} = e^{1.42}\cdot e^{-0.64s} = 4.14 \cdot 0.53c \\
  p_{ij}|_{c=0} & = \frac{4.14}{1 + 4.14} = 0.81 \\
  p_{ij}|_{c=1} & = \frac{4.14\cdot0.53}{1 + 4.14\cdot{0.53}} = 0.69
\end{align}
%
Turning on the balance-assist system reduces the probability that the
perturbation results in a fall from 81\% to 69\%.

This illustration alludes to the difficulty needed to apply the model in way
that could predict how many falls may be averted in a natural setting if the
balance assist system is used. Estimates of the predictor variables extracted
from the limited data collected from natural cycling would be needed to
populate the model.

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why the balance assist system did not have a statistically
significant effect at 10~\si{\kph}. As a bicycle travels at higher speeds, the
same perturbation magnitude causes larger lateral deviations. At 10~\si{\kph}
almost all falls were due to the bicycle exiting the maximum width of the
treadmill. If the same experiment was performed on an infinitely wide plane,
the riders may have recovered from more perturbations. At 6~\si{\kph} the
riders could often recover in the allotted treadmill width due to the smaller
lateral deviations. We believe our results are very much dependent on the two
modes of falling with use: exit the treadmill width or foot is placed on the
belt. Cycle paths are a similar width as the treadmill, so rider's are often
limited in width when recovering from a fall.

@@

Roll and steer angle are not a significant predictor of fall probability at 6
or at 10~\si{\kph}. We expected these variables to have a significant effect
based on the following reasoning. If you are in a rolled and steered state that
is far from the upright equilibrium, then a perturbation that further pushes
you from the equilibrium should have some additive or multiplicative effect on
the resulting motion trajectory.

@@

\section{Conclusion}
%
Automatically controlling a steering motor in a bicycle using roll rate
feedback lowers the speed at which the bicycle is stable. This makes the
bicycle's low speed dynamics more akin to its uncontrolled high speed dynamics,
which is easier for a rider to balance. Perturbation forces applied to the
handlebar can cause a rider to fall and every rider has a threshold force at
which they are more likely to fall than not. The probability of falling when
mechanically perturbed is significantly reduced when traveling at 6~\si{\kph}
when the balance assisting control is activated. This effect is present when
traveling at 10~\si{\kph} but more investigation is needed to determine if the
effect can be significant. The positive effect to balance is rider independent
and most effective in the regime of perturbations near the rider's control
authority threshold. Given that similar effects cause falls during bicycling,
use of the balance assist system in real world use cases will reduce the number
of falls at low bicycling speeds.

\section*{Acknowledgements}
%
This study follows and draws from experimental and analysis methods originally
developed in unpublished research by Marco Reijne. The authors acknowledge
Felix Dauer, David Gabriel, Sierd Heida, Oliver Maier, Maarten Pelgrim, Marco
Reijne, and Arend Schwab for contributions to the development of the balance
assist bicycle.

\section*{Funding}
%
This study is funded by Dutch Research Council, Nederlandse Organisatie voor
Wetenschappelijk Onderzoek (NWO), under the Citius Altius Sanius program and in
collaboration with Bosch eBike Systems and Royal Dutch Gazelle. The funders had
no role in the data collection and analysis or preparation of the manuscript.
